---
title: "Introduction to geomod3D"
author: "Italo Gomes Goncalves"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to geomod3D}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```

Welcome to the `geomod3D` package. This package was designed to allow easy
visualization and manipulation of 3D geospatial data, as well as an open source
alternative to implicit 3D geological modeling. This vignette contains a simple
example of manipulation and modeling of drillhole data from the Ararangua 
region in southern Brazil.

## Importing drillhole data

There are two kinds of data that are needed to generate the 3D object: collar 
and assay/lithological data. The data looks like this:

```{r data}
library(geomod3D)
library(rgl)

knitr::kable(ara.collar)
knitr::kable(head(ara.lito))
```

Let's use the example data to generate a `lines3DDataFrame` object:

```{r lines3D}
ara.dh <- lines3DDataFrame(collar = ara.collar, assay = ara.lito)
ara.dh
```

If needed, the coordinates and variables can be recovered with getter functions.

```{r get}
knitr::kable(head(GetCoords(ara.dh, as = "data.frame")))
knitr::kable(head(GetData(ara.dh)))
```

### Subsetting

Any object that extends the `spatial3DDataFrame` class can be subsetted much 
like a data frame, with the `[` operator. The only recommendation is to use 
column names instead of numbers.

```{r subsetting}
ara.dh[1:5, ]
ara.dh[, c("HOLEID", "Form")]
```

A single column can be extracted directly with the `[[` operator:

```{r subsetting2}
unique(ara.dh[["Form"]])
```

Assignment can also be done the same way as in a data frame:

```{r assign}
ara.dh[, "Length.feet"] <- 0.3 * ara.dh[["Length"]]
```


### Visualization

The next step is visualization. The `Draw*` functions add graphical objects to
the current `rgl` window. If there is none, one will be created.

```{r vis}
# Color coding for the geological formations
formcolor <- c("green","cornflowerblue","burlywood4","firebrick",
               "darkorchid","khaki")
formval <- c("Recente", "Rio do Rasto", "Estrada Nova", "Irati",
             "Palermo", "Rio Bonito")

# Opening rgl window to display in html page
invisible(open3d(useNULL = T, windowRect = c(0,0,800,800), zoom = 1))

# Drawing holes
DrawDrillholes(ara.dh, by = "Form", values = formval, col = formcolor,
               size = 100)

# Hole ID
DrawHoleID(ara.dh)

# The aspect3d function can be used to control the vertical exaggeration
aspect3d(1.731788, 1, 0.2)

# Axes
axes3d()

# Showing scene
rglwidget(width = 800, height = 600)
```

## Modeling

In order to make an implicit model, first the data is converted to a suitable
format, then we choose a set of parameters and build a Gaussian Process model.

### Preparing the data

The drillhole data is processed to extract points inside each geological 
formation and also in the boundaries between formations. The package contains 
a series of methods to perform this task.

As we are modeling the geological formation, the other variables can be 
discarded and adjacent segments of the same formation can be merged in 
order to reduce the size of the dataset.

```{r merge}
ara.dh.proc <- MergeSegments(ara.dh, by = "Form", keep = "Form")
ara.dh.proc
```

Next we extract points containing the geological information. As a point can 
belong to a geological contact, each point carries two labels. If the labels 
differ, then the point lies in a contact.

```{r pointify}
# Duplication of labels for the line segments
ara.dh.proc[, "Form.down"] <- ara.dh.proc[, "Form.up"] <- ara.dh.proc[, "Form"]

# Conversion to point data
ara.point <- Pointify(ara.dh.proc, locations = seq(0.1, 0.9, 0.2))
ara.point

# Finding the contacts
ara.contacts <- GetContacts(ara.dh.proc, by = "Form")
ara.contacts

# Joining the two objects
ara.point <- Bind(ara.point, ara.contacts)
ara.point
```


### The implicit model

The first step in implciit modeling is the definition of a covariance model. As
the geological layers seem to be stratified, the range in the vertical 
direction is made smaller than in the horizontal:

```{r cov_model}
cov_model <- covarianceStructure3D(type = "cubic", 
                                   contribution = 0.1,
                                   maxrange = 5000,
                                   midrange = 5000,
                                   minrange = 200)
```

The parameters were arbitrarily chosen for this example, but the model will be
properly trained later. Next we define a point grid over which to make the
predictions:

```{r grid}
point.grid <- grid3DDataFrame(gridx = seq(647000, 661000, 200),
                              gridy = seq(6791000, 6799000, 200),
                              gridz = seq(-550, 50, 25), 
                              fields = "Form")
```

Then we build a Gaussian Process model and use it to make predictions:

```{r GP, cache=TRUE}
gp.model <- GP_geomod(data = ara.point, 
                      value1 = "Form.up", value2 = "Form.down",
                      model = cov_model, nugget = 0.02)
point.grid <- Predict(gp.model, target = point.grid, to = "Form")
point.grid
```

### Model visualization

The geological contacts can be drawn by transforming each indicator variable 
into a 3D array and feeding it to the `misc3d::contour3d()` function. Note 
that the model matches the geological contacts in the data.

```{r model_vis}
library(misc3d)

# Opening rgl window to display in html page
invisible(open3d(useNULL = T, windowRect = c(0,0,800,800), zoom = 1))

# Drawing holes
DrawDrillholes(ara.dh, by = "Form", values = formval, col = formcolor,
               size = 100)

# Contouring each indicator variable
form <- Make3DArray(point.grid, "Form..Recente.ind")
display1 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "green", alpha = 1, add = T)
form <- Make3DArray(point.grid, "Form..Rio.do.Rasto.ind")
display2 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "cornflowerblue", alpha = 1, add = T)
form <- Make3DArray(point.grid, "Form..Estrada.Nova.ind")
display3 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "burlywood4", alpha = 1, add = T)
form <- Make3DArray(point.grid, "Form..Irati.ind")
display4 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "firebrick", alpha = 1, add = T)
form <- Make3DArray(point.grid, "Form..Palermo.ind")
display5 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "darkorchid", alpha = 1, add = T)
form <- Make3DArray(point.grid, "Form..Rio.Bonito.ind")
display6 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "khaki", alpha = 1, add = T)

# Aspect ratio and axes
aspect3d(1.731788, 1, 0.2)
axes3d()

# Display scene
rglwidget(width = 800, height = 600) %>%
  toggleWidget(ids = display1, label = "Recente") %>%
  toggleWidget(ids = display2, label = "Rio do Rasto") %>%
  toggleWidget(ids = display3, label = "Estrada Nova") %>%
  toggleWidget(ids = display4, label = "Irati") %>%
  toggleWidget(ids = display5, label = "Palermo") %>%
  toggleWidget(ids = display6, label = "Rio Bonito")
```

The geological layers appear closed away from the data. This happens because 
the model steadily loses confidence as we move away from the data points. 
Grid nodes in those regions receive the label `"Unknown"`.

### Model training

The model above was generated with arbitrary parameters. It uses the 
log-likelihood as a goodness-of-fit measure:

```{r loglik1}
logLik(gp.model)
```

The `GP_geomod` class features a genetic algorithm that tries to improve the 
log-likelihood by adjusting the parameters. This is done through the `Fit()` 
method:

```{r training, cache=TRUE}
set.seed(1234) # for reproducibility
gp.model.fitted <- Fit(gp.model, maxrange = T, minrange = T)
logLik(gp.model.fitted)
```

We see that the new model fits the data better. The optimized parameters are:

```{r opt_par}
gp.model.fitted@GPs[[1]]@model
gp.model.fitted@params$nugget
```

The new model can be checked by repeating the commands in the previous 
sections:

```{r GP2, cache=TRUE, echo=FALSE}
point.grid <- Predict(gp.model.fitted, target = point.grid, to = "Form_fitted")
```

```{r model_vis2, echo=FALSE}
# Opening rgl window to display in html page
invisible(open3d(useNULL = T, windowRect = c(0,0,800,800), zoom = 1))

# Drawing holes
DrawDrillholes(ara.dh, by = "Form", values = formval, col = formcolor,
               size = 100)

# Contouring each indicator variable
form <- Make3DArray(point.grid, "Form_fitted..Recente.ind")
display1 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "green", alpha = 1, add = T)
form <- Make3DArray(point.grid, "Form_fitted..Rio.do.Rasto.ind")
display2 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "cornflowerblue", alpha = 1, add = T)
form <- Make3DArray(point.grid, "Form_fitted..Estrada.Nova.ind")
display3 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "burlywood4", alpha = 1, add = T)
form <- Make3DArray(point.grid, "Form_fitted..Irati.ind")
display4 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "firebrick", alpha = 1, add = T)
form <- Make3DArray(point.grid, "Form_fitted..Palermo.ind")
display5 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "darkorchid", alpha = 1, add = T)
form <- Make3DArray(point.grid, "Form_fitted..Rio.Bonito.ind")
display6 <- contour3d(form$value, level = 0, 
                      x = form$x, y = form$y, z = form$z, 
                      color = "khaki", alpha = 1, add = T)

# Aspect ratio and axes
aspect3d(1.731788, 1, 0.2)
axes3d()

# Display scene
rglwidget(width = 800, height = 600) %>%
  toggleWidget(ids = display1, label = "Recente") %>%
  toggleWidget(ids = display2, label = "Rio do Rasto") %>%
  toggleWidget(ids = display3, label = "Estrada Nova") %>%
  toggleWidget(ids = display4, label = "Irati") %>%
  toggleWidget(ids = display5, label = "Palermo") %>%
  toggleWidget(ids = display6, label = "Rio Bonito")
```
